---
title: "Preventing Whale Strikes in the Santa Barbara Channel"
author: "Laura Ingulsrud"
date: "5/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}

# Load necessary packages
library(tidyverse)

```

```{r data, include=FALSE}

# Load in data set from whale valuation surveys
whalesurveydata <- read_csv("HW3Data.csv")

# Data columns are: 
  # (1) “risk”: level of risk reduction
  # (2) “bid”: annual payment for the household
  # (3) “vote”: 1 is yes, 0 is no
  # (4) "NEP”: measure of environmental concern
  # (5) "income”: categorical variable for income level 
  # (6) "age”: categorical variable for age

```

###**1.** Linear Probability Model
```{r #1, include=FALSE, echo=FALSE}

# 1. Create a linear probability model that predicts a respondent’s probability of voting ‘yes’ on the ballot based on their age, income, NEP score, the risk reduction offered by the program, and the cost of the program to that respondent. Show the model and interpret the regression coefficients.

# Do a linear regression through the data
whalesurvey_lm <- lm(vote ~ risk + bid + vote + NEP + income + age, data = whalesurveydata)

summary(whalesurvey_lm)

# Assign coefficients
intercept <- whalesurvey_lm$coefficients[1]
risk <- whalesurvey_lm$coefficients[2]
bid <- whalesurvey_lm$coefficients[3]
NEP <- whalesurvey_lm$coefficients[4]
one_percent <- whalesurvey_lm$coefficients[5]
poor <- whalesurvey_lm$coefficients[6]
rich <- whalesurvey_lm$coefficients[7]
very_rich <- whalesurvey_lm$coefficients[8]
ageto50 <- whalesurvey_lm$coefficients[9]
ageto40 <- whalesurvey_lm$coefficients[10]
ageto60 <- whalesurvey_lm$coefficients[11]
ageto30 <- whalesurvey_lm$coefficients[12]

# Reference coefficients:
  # Age: oversixty
  # Income: middle

```
$Probability~of~Voting~Yes = (`r round(intercept, digits = 4)`) + (`r round(ageto30, digits = 4)`)(age~to~thirty) + (`r round(ageto40, digits = 4)`)(age~to~forty) + (`r round(ageto50, digits = 4)`)(age~to~fifty) + (`r round(ageto60, digits = 4)`)(age~to~sixty) + (`r round(poor, digits = 4)`)(poor~income~level) + (`r round(rich, digits = 4)`)(rich~income~level) + (`r round(very_rich,digits = 4)`)(very~rich~income~level) +(`r round(one_percent, digits = 4)`)(one~percent~income~level) +  (`r round(NEP, digits = 4)`)(NEP) + (`r round(bid, digits = 4)`)(bid) + (`r round(risk, digits = 4)`)(risk~reduction~level)$

<br>

With all other variables held constant, respondents in the age to thirty group are 2.04% more likely to vote yes than respondents in the over sixty age group. <br>
With all other variables held constant, respondents in the age to forty group are 2.01% less likely to vote yes than respondents in the over sixty age group. <br>
With all other variables held constant, respondents in the age to fifty group are 0.10% more likely to vote yes than respondents in the over sixty age group. <br>
With all other variables held constant, respondents in the age to sixty group are 1.62% less likely to vote yes than respondents in the over sixty age group. <br>
With all other variables held constant, respondents in the poor income level are 0.27% more likely to vote yes than respondents in the middle income level. <br>
With all other variables held constant, respondents in the rich income level are 0.75% more likely to vote yes than respondents in the middle income level. <br>
With all other variables held constant, respondents in the very rich income level are 4.68% more likely to vote yes than respondents in the middle income level. <br>
With all other variables held constant, respondents in the one percent income level are 0.88% more likely to vote yes than respondents in the middle income level. <br>
With all other variables held constant, for every 1 point increase in NEP (measure of environmental concern), there is a 1.59% greater chance of voting yes. <br>
With all other variables held constant, for every $1 increase in bid (annual payment for household), there is a 0.11% less chance of voting yes. <br>
With all other variables held constant, for every 1% increase in risk reduction level, there is a 0.07% greater chance of voting yes. <br>

<br>
<br>

###**2.** 
```{r #2}

# 2. Based on this regression, what can you say about the value of a single prevented whale death? (Hint: Think about how risk reduction for all whales translates into the number of whale deaths avoided)


```



```{r #3}

# 3. Pick three arbitrary respondents. For each of those respondents, estimate their WTP for a VSR program offering 60% risk reduction.

```

```{r #4}

# 4. Now repeat this process and estimate the mean WTP for a VSR program offering 60% risk reduction among Santa Barbara County households.

```

```{r #5}

# 5. If there are 150,000 households in Santa Barbara County, estimate the total benefit of a VSR program to these households.

```

```{r #6}

# 6. Based only on these benefits, do the benefits of the VSR program outweigh the costs?

```

```{r #7}

# 7. Suppose the price of a carbon credit is $Z per ton, at what price Z* will the shipping industry decide to adopt the VSR for purely self-interested reasons? (Note: if ships voluntarily reduce speed, then the $7 million cost (noted above) of implementing the VSR would be avoided)

```

```{r #8}

# 8. Now suppose the carbon credit price is exactly Z*, so all ships voluntarily reduce speed to achieve the 60% risk reduction. Approximately how many whales would be saved by this? What would be the social value of allowing ships to enter the carbon trading market?

```
