---
output:
  html_document:
    df_print: paged
    geometry: margin=0cm
    theme: cosmo
    fontsize: 12pt
  pdf_document: default
---

<style type="text/css">
.main-container {
  max-width: 1800px;
  margin-left: auto;
  margin-right: auto;
}
</style>

**To:** Santa Barbara County Board of Supervisors <br>
**From:** Laura Ingulsrud & Jenny Balmagia, WorthWhale Consulting LLC <br>
**Re:** Economic Benefits of Reducing Whale Strikes in the Santa Barbara Channel <br>
**Date:** 23 May 2019 <br><br>

**Introduction** <br>
In our analysis of whether Santa Barbara County (County) should implement a vessel speed reduction (VSR) program in the Santa Barbara Channel (Channel) to reduce whale strikes, we  quantified the costs and benefits and found the program to be a good investment. To reach this conclusion, we determined the County resident’s demand, or willingness-to-pay (WTP), for fewer whale strikes using a contingent valuation model (CVM) and compared this to the cost of implementation.

**Analysis and Results** <br>
This analysis estimates the expected benefits of a VSR program through a CVM survey distributed to 500 County residents. Using the survey results, we extrapolated County resident’s stated preference or WTP for reducing whale strikes in the Channel. This was accomplished by creating a linear regression model in R with the survey data predicting an individual respondent’s probability of voting “yes” for the VSR program.  We used this model to predict the median WTP for a 60% reduction in whale strikes, or 15 whales saved annually. From this, we found County residents (150,000 households) are willing to spend $42,378,337 to prevent 15 whale deaths. Considering the cost of implementing the VSR program is $7 million, residents would experience a benefit of $35,378,337 if the program were funded by taxpayers.  

**Recommendations** <br>
Given these results, the Board of Supervisors should encourage implementation of the VSR program, as the benefits to County residents far outweigh the costs. As stated above, if residents are taxed a total of $7 million to implement the VSR program, they experience a benefit of $35,378,337. However, with a carbon trading market that offers carbon credits of at least $50/ton of CO2, shippers will voluntarily opt into the VSR program and eliminate the $7 million implementation cost. By reducing their speed by 60%, ships will not only reduce their probability of striking a whale but emit less CO2. At a carbon credit price of $50/ton of CO2, shippers would regain any revenue lost by slowing down. If the shipping industry were to participate in such a carbon offset program, the net benefit to residents would increase to $42,378,337 as the $7 million cost would be avoided. Adoption of either scenario will result in a significant positive net benefit for County residents, with the carbon trading market scenario resulting in the greatest net benefit. As this program would save 15 whales a year from lethal ship strikes, County residents will benefit from the continued presence of these charismatic species for generations to come.

<P style="page-break-before: always">

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}

# Load necessary packages
library(tidyverse)

```

```{r data, include=FALSE}

# Load in data set from whale valuation surveys
whalesurveydata <- read_csv("HW3Data.csv")

# Data columns are: 
  # (1) “risk”: level of risk reduction
  # (2) “bid”: annual payment for the household
  # (3) “vote”: 1 is yes, 0 is no
  # (4) "NEP”: measure of environmental concern
  # (5) "income”: categorical variable for income level 
  # (6) "age”: categorical variable for age

```

##**Technical Appendix**<br>

###1. Linear Probability Model
```{r #1, include=FALSE, echo=FALSE}

# 1. Create a linear probability model that predicts a respondent’s probability of voting ‘yes’ on the ballot based on their age, income, NEP score, the risk reduction offered by the program, and the cost of the program to that respondent. Show the model and interpret the regression coefficients.

# Do a linear regression through the data
whalesurvey_lm <- lm(vote ~ risk + bid + vote + NEP + income + age, data = whalesurveydata)

summary(whalesurvey_lm)
# Coefficients:
#                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)        0.1196977  0.1198911   0.998    0.319    
# risk               0.0007445  0.0008363   0.890    0.374    
# bid               -0.0010699  0.0006585  -1.625    0.105    
# NEP                0.0158639  0.0020887   7.595 1.58e-13 ***
# incomeone_percent  0.0088282  0.0598973   0.147    0.883    
# incomepoor         0.0027386  0.0649833   0.042    0.966    
# incomerich         0.0074891  0.0682176   0.110    0.913    
# incomevery_rich    0.0467922  0.0674876   0.693    0.488    
# agetofifty         0.0099816  0.0633105   0.158    0.875    
# agetoforty        -0.0201190  0.0623958  -0.322    0.747    
# agetosixty        -0.0162261  0.0595666  -0.272    0.785    
# agetothirty        0.0204401  0.0578269   0.353    0.724    

# Assign coefficients
intercept <- whalesurvey_lm$coefficients[1]
risk <- whalesurvey_lm$coefficients[2]
bid <- whalesurvey_lm$coefficients[3]
NEP <- whalesurvey_lm$coefficients[4]
one_percent <- whalesurvey_lm$coefficients[5]
poor <- whalesurvey_lm$coefficients[6]
rich <- whalesurvey_lm$coefficients[7]
very_rich <- whalesurvey_lm$coefficients[8]
ageto50 <- whalesurvey_lm$coefficients[9]
ageto40 <- whalesurvey_lm$coefficients[10]
ageto60 <- whalesurvey_lm$coefficients[11]
ageto30 <- whalesurvey_lm$coefficients[12]

# Reference coefficients:
  # Age: oversixty
  # Income: middle

########################### EXTRA CREDIT ###########################
whalesurvey_glm <- glm(vote ~ risk + bid + vote + NEP + income + age, data = whalesurveydata, family = binomial)

summary(whalesurvey_glm)
# Coefficients:
#                    Estimate Std. Error z value Pr(>|z|)    
# (Intercept)       -2.088292   0.665981  -3.136  0.00171 ** 
# risk               0.004169   0.004631   0.900  0.36802    
# bid               -0.005865   0.003639  -1.612  0.10700    
# NEP                0.083213   0.012273   6.780  1.2e-11 ***
# incomeone_percent  0.059366   0.331487   0.179  0.85787    
# incomepoor         0.040298   0.359092   0.112  0.91065    
# incomerich         0.060219   0.376718   0.160  0.87300    
# incomevery_rich    0.277879   0.380863   0.730  0.46563    
# agetofifty         0.062771   0.355469   0.177  0.85983    
# agetoforty        -0.127045   0.336290  -0.378  0.70559    
# agetosixty        -0.107886   0.330639  -0.326  0.74420    
# agetothirty        0.092936   0.318546   0.292  0.77048  


# Assign coefficients (EXTRA CREDIT)
interceptlog <- whalesurvey_glm$coefficients[1]
risklog <- whalesurvey_glm$coefficients[2]
bidlog <- whalesurvey_glm$coefficients[3]
NEPlog <- whalesurvey_glm$coefficients[4]
one_percentlog <- whalesurvey_glm$coefficients[5]
poorlog <- whalesurvey_glm$coefficients[6]
richlog <- whalesurvey_glm$coefficients[7]
very_richlog <- whalesurvey_glm$coefficients[8]
ageto50log <- whalesurvey_glm$coefficients[9]
ageto40log <- whalesurvey_glm$coefficients[10]
ageto60log <- whalesurvey_glm$coefficients[11]
ageto30log <- whalesurvey_glm$coefficients[12]

# Reference coefficients:
  # Age: oversixty
  # Income: middle

#AIC of glm model, 561. AIC of linear model, 586. Not super different which means fit is fairly similar to both, but logically glm should fit better? Linear is overestimating, but why???
```

**Linear Regression Model:** <br>

$Probability~of~Voting~Yes = intercept + age~to~thirty + age~to~forty + age~to~fifty + age~to~sixty + poor~income~level + rich~income~level + very~rich~income~level + one~percent~income~level + NEP + bid + risk~reduction~level$

$Probability~of~Voting~Yes = (`r round(intercept, digits = 4)`) + (`r round(ageto30, digits = 4)`) + (`r round(ageto40, digits = 4)`) + (`r round(ageto50, digits = 4)`) + (`r round(ageto60, digits = 4)`) + (`r round(poor, digits = 4)`) + (`r round(rich, digits = 4)`) + (`r round(very_rich,digits = 4)`) +(`r round(one_percent, digits = 4)`) +  (`r round(NEP, digits = 4)`) + (`r round(bid, digits = 4)`) + (`r round(risk, digits = 4)`)$

<br>
<center>
```{r stargazer table, results="asis", echo=FALSE}

stargazer::stargazer(whalesurvey_lm, type = "html", title = "Linear Regression Coefficients", digits = 4)

########################### EXTRA CREDIT ###########################
stargazer::stargazer(whalesurvey_glm, type = "html", title = "Logistic Regression Coefficients", digits = 4)

```
</center>
<br>

**Reference coefficients:**

- Age: oversixty <br>
- Income: middle <br>

**Regression Coefficient Interpretations:**

- With all other variables held constant, respondents in the age to thirty group are 2.04% more likely to vote yes than respondents in the over sixty age group. <br>
- With all other variables held constant, respondents in the age to forty group are 2.01% less likely to vote yes than respondents in the over sixty age group. <br>
- With all other variables held constant, respondents in the age to fifty group are 0.10% more likely to vote yes than respondents in the over sixty age group. <br>
- With all other variables held constant, respondents in the age to sixty group are 1.62% less likely to vote yes than respondents in the over sixty age group. <br>
- With all other variables held constant, respondents in the poor income level are 0.27% more likely to vote yes than respondents in the middle income level. <br>
- With all other variables held constant, respondents in the rich income level are 0.75% more likely to vote yes than respondents in the middle income level. <br>
- With all other variables held constant, respondents in the very rich income level are 4.68% more likely to vote yes than respondents in the middle income level. <br>
- With all other variables held constant, respondents in the one percent income level are 0.88% more likely to vote yes than respondents in the middle income level. <br>
- With all other variables held constant, for every 1 point increase in NEP (measure of environmental concern), there is a 1.59% greater chance of voting yes. <br>
- With all other variables held constant, for every $1 increase in bid (annual payment for household), there is a 0.11% less chance of voting yes. <br>
- With all other variables held constant, for every 1% increase in risk reduction level, there is a 0.07% greater chance of voting yes. <br><br>

###2. Value of a Single Prevented Whale Death
```{r #2, include=FALSE, echo=FALSE}

# 2. Based on this regression, what can you say about the value of a single prevented whale death? (Hint: Think about how risk reduction for all whales translates into the number of whale deaths avoided)


#Mode for categorical variables
#age
whalesurveyage <- whalesurveydata %>% 
group_by(age) %>% 
  count(age)
#To thirty is most respondents, n = 123

#income 
whalesurveyincome <- whalesurveydata %>% 
group_by(income) %>% 
count(income)
#One-percent is most respondents n = 158 


#Average NEP 
meanNEP <- mean(whalesurveydata$NEP)

#Value of 1 whale 

bid0 <- (0.5 - intercept - risk*0 - NEP*meanNEP - one_percent - ageto30)/bid

bid4 <- (0.5 - intercept - risk*4 - NEP*meanNEP - one_percent - ageto30)/bid

bid_diff <- bid4 - bid0

bid_diff # $2.78

########################### EXTRA CREDIT ###########################

#Value of 1 whale 

bid0log <- (0 - interceptlog - risklog*0 - NEPlog*meanNEP - one_percentlog - ageto30log)/bidlog

bid4log <- (0 - interceptlog - risklog*4 - NEPlog*meanNEP - one_percentlog - ageto30log)/bidlog

bid_difflog <- bid4log - bid0log

bid_difflog # $2.84

```

If a risk reduction of 20% saves 5 whales, then a risk reduction of 4% saves 1 whale. We created two functions, one assuming a risk reduction of 0% and one of 4%, then took the difference in value of these two functions to find the value of a single prevented whale death is **$2.78**. <br>

$Bid(0) = \frac{0.5 - intercept - (risk~reduction~level)(0~percent) - (NEP)(mean~NEP) - one~percent~income~level - age~to~30}{bid} = \frac{0.5 - ((`r round(intercept, digits = 4)`) - (`r round(risk, digits = 4)`)(0) - (`r round(NEP, digits = 4)`)(`r round(meanNEP, digits = 4)`) - (`r round(one_percent, digits = 4)`) - (`r round(ageto30, digits = 4)`))}{`r round(bid, digits = 4)`}$ <br>

$Bid(4) = \frac{0.5 - intercept - (risk~reduction~level)(4~percent) - (NEP)(mean~NEP) - one~percent~income~level - age~to~30}{bid} = \frac{0.5 - ((`r round(intercept, digits = 4)`) - (`r round(risk, digits = 4)`)(4) - (`r round(NEP, digits = 4)`)(`r round(meanNEP, digits = 4)`) - (`r round(one_percent, digits = 4)`) - (`r round(ageto30, digits = 4)`))}{`r round(bid, digits = 4)`}$ <br>

$Bid(4) - Bid(0)$ = $2.78 <br>

**Assumptions:**

- We used the mode for the categorical variables of age and income: The mode for the age variable was "ageto30" and the mode for the income variable was "one_percent". <br>
- We used the mean for the continuous variable, NEP. <br>
- We used 0.5 for the vote probability, since using a value of 0.5 (halfway between 0 and 1), will give us the median WTP. <br>
- *Note: These assumptions apply for #2-8.* <br><br>

###3. Random Respondents' WTP
```{r #3, include=FALSE, echo=FALSE}

# 3. Pick three arbitrary respondents. For each of those respondents, estimate their WTP for a VSR program offering 60% risk reduction.

#Person 1

bid1 <- (0.5 - intercept - risk*60 - NEP*44 - very_rich - ageto60)/bid
bid1 # $367.27

bid2 <- (0.5 - intercept - risk*60 - NEP*39 - poor - ageto30)/bid
bid2 # $286.23

bid3 <- (0.5 - intercept - risk*60 - NEP*53 - one_percent - ageto50)/bid
bid3 # $489.73

########################### EXTRA CREDIT ###########################

bid1log <- (0 - interceptlog - risklog*60 - NEPlog*44 - very_richlog - ageto60log)/bidlog
bid1log # $339.84

bid2log <- (0 - interceptlog - risklog*60 - NEPlog*39 - poorlog - ageto30log)/bidlog
bid2log # $262.63

bid3log <- (0 - interceptlog - risklog*60 - NEPlog*53 - one_percentlog - ageto50log)/bidlog
bid3log # $459.37

```

We used the same equation from #2, but instead of using the means and modes for the variables, we plugged in these numbers for each random respondent. <br>

*Respondent 1:* <br>
- NEP = 44 <br>
- income = very_rich <br>
- age = ageto60 <br>
- **WTP = $367.27** <br>
  
$Bid~Average_1 = \frac{0.5 - intercept - (risk~reduction~level)(60~percent) - (NEP)(44) - very~rich~income~level - age~to~60}{bid} = \frac{0.5 - ((`r round(intercept, digits = 4)`) - (`r round(risk, digits = 4)`)(60) - (`r round(NEP, digits = 4)`)(44) - (`r round(very_rich, digits = 4)`) - (`r round(ageto60, digits = 4)`))}{`r round(bid, digits = 4)`} = 367.27$ <br><br>

*Respondent 2:* <br>
- NEP = 39 <br>
- income = poor <br>
- age = ageto30 <br>
- **WTP = $286.23** <br>
  
$Bid~Average_2 = \frac{0.5 - intercept - (risk~reduction~level)(60~percent) - (NEP)(39) - one~percent~income~level - age~to~50}{bid} = \frac{0.5 - ((`r round(intercept, digits = 4)`) - (`r round(risk, digits = 4)`)(60) - (`r round(NEP, digits = 4)`)(39) - (`r round(one_percent, digits = 4)`) - (`r round(ageto50, digits = 4)`))}{`r round(bid, digits = 4)`} = 286.23$ <br><br>
  
*Respondent 3:* <br>
- NEP = 53 <br>
- income = one_percent <br>
- age = ageto50 <br>
- **WTP = $489.73** <br>

$Bid~Average_3 = \frac{0.5 - intercept - (risk~reduction~level)(60~percent) - (NEP)(53) - poor~income~level - age~to~30}{bid} = \frac{0.5 - ((`r round(intercept, digits = 4)`) - (`r round(risk, digits = 4)`)(60) - (`r round(NEP, digits = 4)`)(53) - (`r round(poor, digits = 4)`) - (`r round(ageto30, digits = 4)`))}{`r round(bid, digits = 4)`} = 489.73$ <br>

<br>

###4. Mean WTP for VSR Program
```{r #4, include=FALSE, echo=FALSE}

# 4. Now repeat this process and estimate the mean WTP for a VSR program offering 60% risk reduction among Santa Barbara County households.

bid_average <- (0.5 - intercept - risk*60 - NEP*meanNEP - one_percent - ageto30)/bid
bid_average # $282.52

########################### EXTRA CREDIT ###########################

bid_averagelog <- (0 - interceptlog - risklog*60 - NEPlog*meanNEP - one_percentlog - ageto30log)/bidlog
bid_averagelog # $256.89

```

We used the same assumptions as #2 and a risk reduction of 60% to get a mean WTP of **$282.52**. <br>

$Bid~Average = \frac{0.5 - intercept - (risk~reduction~level)(60~percent) - (NEP)(mean~NEP) - one~percent~income~level - age~to~30}{bid} = \frac{0.5 - ((`r round(intercept, digits = 4)`) - (`r round(risk, digits = 4)`)(60) - (`r round(NEP, digits = 4)`)(`r round(meanNEP, digits = 4)`) - (`r round(one_percent, digits = 4)`) - (`r round(ageto30, digits = 4)`))}{`r round(bid, digits = 4)`} = 282.52$ <br><br>

###5. Total Benefit of VSR Program to Santa Barbara County Households
```{r #5, include=FALSE, echo=FALSE}

# 5. If there are 150,000 households in Santa Barbara County, estimate the total benefit of a VSR program to these households.

total_benefit <- 150000*bid_average
total_benefit # $42,378,337

########################### EXTRA CREDIT ###########################

total_benefitlog <- 150000*bid_averagelog
total_benefitlog # $38,534,051

```

We multiplied 150,000 households by the mean WTP ($282.52) to find a total benefit of **$42,378,337**. Because you can assume Santa Barbara County residents are willing to pay this amount for having these whales exist, that amount equates to the total benefit residents are receiving. <br><br>

###6. VSR Program Benefit-Cost Comparison
```{r #6, include=FALSE, echo=FALSE}

# 6. Based only on these benefits, do the benefits of the VSR program outweigh the costs?

#Yes, by $35,378,337
net_benefits <- total_benefit - 7000000
net_benefits

########################### EXTRA CREDIT ###########################

#Yes, by $31,534,051
net_benefitslog <- total_benefitlog - 7000000
net_benefitslog

```

We subtracted the $7 million cost of the VSR program from the total benefit of $42,378,337 to find that **yes, the benefits of the VSR program outweigh the costs by $35,378,337.** <br><br>

###7. Carbon Credit Price Needed for Shipping Industry to Adopt VSR Program
```{r #7, include=FALSE, echo=FALSE}

# 7. Suppose the price of a carbon credit is $Z per ton, at what price Z* will the shipping industry decide to adopt the VSR for purely self-interested reasons? (Note: if ships voluntarily reduce speed, then the $7 million cost (noted above) of implementing the VSR would be avoided)

#Find the price of what 1 ton of carbon would have to be given that saving 20tons costs shippers $1000
1000/20

#$50/ton

```

If ships voluntarily reduce speed, the cost to each shipper would be $1,000 and 20 tons of CO2 emissions per transit would be reduced. Therefore, the price of 1 ton of CO2 is worth $50 ($1,000/20 tons). If there's a carbon credit worth **$50/ton of CO2**, the shipping industry wouldn't lose any money, and would adopt the VSR for purely self-interested reasons. <br><br>

###8. Whales Saved by VSR Program Implementation Through Carbon Trading Markets
```{r #8, include=FALSE, echo=FALSE}

# 8. Now suppose the carbon credit price is exactly Z*, so all ships voluntarily reduce speed to achieve the 60% risk reduction. Approximately how many whales would be saved by this? What would be the social value of allowing ships to enter the carbon trading market?

#Social Value - same as 5 since value of offset carbon and compliance cost cancel (they're the same)

#Number of whales saved 
#If 20% reduction saves 5 whales, then 60% reduction saves 15 whales


```

If a 20% risk reduction saves 5 whales, then a 60% risk reduction saves **15 whales**. <br>

*social value = value of program + value of offset carbon - compliance cost* <br>

Sine the value of offset carbon and the compliance cost are the same, they cancel each other out. Therefore, the social value of allowing ships to enter the carbon trading market is the same as the value of the program, which we found to be **$42,378,337** (#5).